\Title

### 2023-06-15 Grimaldi-etal-BiolCybernetics.tex
correc Miles: Results

By pruning the weights, we observe that the evolution of accuracy as a function of the log percentage of active weights fits well to a sigmoid curve. Half saturation is reached at a ratio of about $3.2\ times 10^{-2}$ of active weights, which in our setting corresponds to a total of about $180$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained relative to its peak performance when the number of computations is divided by a factor up to about $31$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, demonstrating the efficiency of the presented method.
%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/quant_accuracy.pdf}
    \caption{
        Accuracy as a function of computational load for the HD-SNN model (blue dots) with error bars indicating the 5\% - 95\% quantiles and a sigmoid fit (blue line). The relative computational load (on a logarithmic axis) is controlled by changing the percentage of nonzero weights relative to the dense convolution kernel. As we prune the coefficients, we observe a stable accuracy value, with a drop observed at about 25 times fewer computations.
        }
    \label{fig:accuracy}
\end{figure}
%
\section{Discussion}
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN with heterogeneous delays and shown how it compares favorably with state-of-the-art event-based classification algorithms for a visual motion detection task. The learned model has a number of similarities with neurobiological anatomical observations as well as behavioral results. The event-based computations of our method can be drastically reduced by pruning synapses, while maintaining top classification performance. This shows that we can use the precise timing of spikes to improve the performance of neural computations.
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
The HD-SNN model trained and evaluated on a naturalistic motion detection task. The model was defined to provide optimal detection of event-driven spatiotemporal motifs. We have shown that the model, when trained on a dataset of natural images with realistic eye movements, learns kernels similar to those found in early visual cortex~\cite{deangelis_functional_1999, kremkow_push-pull_2016}. We have evaluated the computational cost of this model when implemented in a setting similar to event-based hardware. We show that the use of heterogeneous delays may be an efficient computational solution for future neuromorphic hardware, but also a key to understanding why spikes are a universal component of neural information processing.

We would like to highlight a few innovations in the contributions presented in this paper. First, while~\citep{ghosh_spatiotemporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatiotemporal motifs. Moreover, compared to a representation using this algorithm, the parameters of the model (weights and delays) are explicable, as they directly inform about the evidence  of detection for each spatiotemporal spike motif, where we define \emph{evidence} as the logit of the probability, that is, the inverse sigmoid of the probability. Another novelty is that the model learns the weights and delays simultaneously. For example, the polychronization model~\citep{izhikevich_polychronization_2006} learns only the weights using STDP, while the delays are randomly drawn and their values are frozen during learning. In addition, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified toy problems~\citep{gutig_tempotron_2006}. Another major contribution is to provide a model that is suitable for learning any kind of spatiotemporal spiking motif and that can be trained in a supervised manner by providing a dataset of supervised pairs. Instead of relying on a careful description of the physical rules governing a task, e.g. the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}, this allows a more flexible definition of the model using this properly labeled dataset.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified a number of limitations of our model, which we will now discuss in detail. First, the entire framework is based on a discrete binning of time, which is not compatible with the continuous nature of biological time. We used this binning to efficiently implement the framework on conventional hardware, especially GPUs, to be able to use fast three-dimensional convolutions. We have tested the effect of the width of the time bin and observed it has essentially no effect on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where accuracy was unaffected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes. This mechanism may be implemented by the filtering implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circumvent the need for time discretization by the use of a purely event-based scheme. In fact, it is not necessary to compute the voltage traces between two spikes~\citep{hanuschkin_general_2010}. Thus, it is possible to define a purely event-based framework. Such an architecture could provide promising computational speedups.

Another limitation is that the model is purely feed-forward. Thus, the spikes generated by the postsynaptic neurons are based solely on the information contained in the classical receptive field. However, it is known that neurons in the same layer can interact with each other through lateral interactions, for example in V1, and that this can be the basis for computational principles~\citep{chavane_revisiting_2022}. For example, the combination of neighboring orientations may contribute to image categorization~\citep{perrinet_edge_2015}. Furthermore, neural information is modulated by feedback information, e.g. to distinguish a figure from its background~\citep{roelfsema_early_2016}. Feedback has been shown to be essential for building realistic models of primary visual areas~\citep{boutin_sparse_2020, boutin_effect_2020}, especially to explain non-linear mechanisms~\citep{boutin_pooling_2022}. Currently, mainly due to our use of convolutions, it is not possible to implement these recurrent connections in our implementation (lateral or feedback). However, by inserting new spikes into the list of spikes reaching presynaptic addresses, the generic model is able to incorporate them. While this is theoretically possible, it must be properly tuned in practice so that these recurrent connections do not amplify neuronal activity outside of a homeostatic state (by extinction or explosion).

Such recurrent activity would be essential for the implementation of predictive or anticipatory processes. This is essential in a neural system because it contains several different delays that require temporal alignment~\citep{hogendoorn_predictive_2019}. This has been modeled before to explain, for example, the flash-lag illusion~\citep{khoei_flash-lag_2017}. As mentioned previously, this could be implemented using generalized coordinates (i.e., variables such as position complemented by motion, acceleration, jerk,~\ldots), and ``neurobiologically, using delay operators just means changing synaptic connection strengths to take different mixtures of generalized sensations and their prediction errors''~\citep{perrinet_active_2014}. Our proposed model using heterogeneous delays provides an alternative and elegant implementation solution to this problem.
%
\subsection{Perspectives}
%%%-----------------------------------------------------------------
%
In defining our task, we emphasized that the generation of events depends on the spatial gradient in each image. This gradient has both horizontal and vertical dimensions, and its maxima are generally orientation dependent. Taken together, these oriented edges form the contours of visual objects in the scene~\citep{koenderink_representation_1987}. Thus, there is an interdependence between motion information and orientation information within the event stream. It would be crucial to investigate this dependency further. This could be initiated by training the model on a dataset with labels that provide local orientation. Exploring this dependence will allow us to dissociate and integrate these two forms of visual information. In particular, it will allow us to consider that the definition of motion is more accurate perpendicular to an oriented contour (aka the aperture problem). Thus, it will allow us to implement recurrent prediction rules, such as those identified to dissociate this problem~\citep{perrinet_motion-based_2012}.

The model is trained on a low-level local motion detection task, and one might wonder if it could be trained on higher-level tasks. An example of such a task would be depth estimation in the visual scene. There are several sources of information for depth estimation, such as binocular disparity or changes in texture or shading, but in our case motion parallax would be the most important cue~\citep{rogers_motion_1979}. This is because objects that are close to an observer move relatively faster on the retina than an object that is far away, and also because visual occlusions depend on depth order. Using this information, it is possible to segment objects and estimate their depth~\citep{yoonessi_contribution_2011}. However, this would first require the computation of the optic flow, i.e., the extension of the framework described here for a rigid full-field motion to a more general one where the motion may vary in the visual field. One possible implementation is to add a new layer to our model, analogous to the hierarchical organization highlighted in the visual cortex. This is theoretically possible by using the output of our model (which estimates motion in retinotopic space) as input to a new layer of neurons that would estimate motion in the visual field, including the depth dimension in the output supervision labels. This could have direct and important applications, e.g. in autonomous driving to detect obstacles in a fast and robust way. Another extension would be to actively generate sensor motion (physical or virtual) to obtain better depth estimates, especially to disambiguate uncertain estimates~\citep{nawrot_eye_2003}.

>>>>

By pruning the weights, we observe that the evolution of accuracy as a function of the log percentage of active weights fits well to a sigmoid curve. Half saturation is reached at a ratio of about $3.2\ times 10^{-2}$ of active weights, which in our setting corresponds to a total of about $180$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained relative to its peak performance when the number of computations is divided by a factor up to about $31$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, demonstrating the efficiency of the presented method.
%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/quant_accuracy.pdf}
    \caption{
        Accuracy as a function of computational load for the HD-SNN model (blue dots) with error bars indicating the 5\% - 95\% quantiles and a sigmoid fit (blue line). The relative computational load (on a logarithmic axis) is controlled by changing the percentage of nonzero weights relative to the dense convolution kernel. As we prune the coefficients, we observe a stable accuracy value, with a drop observed at about 25 times fewer computations.
        }
    \label{fig:accuracy}
\end{figure}
%
\section{Discussion}
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN with heterogeneous delays and shown how it compares favorably with state-of-the-art event-based classification algorithms for a visual motion detection task. The learned model has a number of similarities with neurobiological anatomical observations as well as behavioral results. The event-based computations of our method can be drastically reduced by pruning synapses, while maintaining top classification performance. This shows that we can use the precise timing of spikes to improve the performance of neural computations.
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
The HD-SNN model was trained and evaluated on a naturalistic motion detection task with realistic eye movements. The model was defined to provide optimal detection of event-driven spatiotemporal motifs and learns kernels similar to those found in the visual cortex~\cite{deangelis_functional_1999, kremkow_push-pull_2016}. We have evaluated the computational cost of this model when implemented in a setting similar to event-based hardware. We show that the use of heterogeneous delays may be an efficient computational solution for future neuromorphic hardware, but also a key to understanding why spikes are a universal component of neural information processing.

We would like to highlight a few innovations in the contributions presented in this paper. First, while~\citep{ghosh_spatiotemporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatiotemporal motifs. Moreover, compared to a representation using this algorithm, the parameters of the model (weights and delays) are explicable, as they directly inform about the evidence  of detection for each spatiotemporal spike motif, where we define \emph{evidence} as the logit of the probability, that is, the inverse sigmoid of the probability. Another novelty is that the model learns simultaneously weights and delays. In contrast, the polychronization model~\citep{izhikevich_polychronization_2006} learns only the weights using STDP, while the delays are randomly drawn and their values are frozen during learning. In addition, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified toy problems~\citep{gutig_tempotron_2006}. Another major contribution is to provide a model that is suitable for learning any kind of spatiotemporal spiking motif and that can be trained in a supervised manner by providing a dataset of supervised pairs. Instead of relying on a careful description of the physical rules governing a task, e.g. the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}, this allows a more flexible definition of the model using this properly labeled dataset.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified a number of limitations of our model, which we will now discuss in detail. First, this implementation of the HD-SNN model is based on a discrete binning of time, which is not compatible with the continuous nature of biological time. We used this binning to efficiently implement the framework on conventional hardware, especially GPUs, to be able to use fast three-dimensional convolutions. We have tested the effect of the width of the time bin and observed it has essentially no effect on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where accuracy was unaffected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the advantage of analytically including a precision term in the temporal value of the input spikes. This mechanism may be implemented by the filtering implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circumvent the need for time discretization by the use of a purely event-based scheme. In fact, it is possible to derive event-triggered computations of the continuous activity of the SNN~\citep{hanuschkin_general_2010} and thus to define a purely event-based framework. Such an architecture could provide promising computational speedups.

Another limitation is that the model is purely feed-forward. Thus, the spikes generated by the postsynaptic neurons are based solely on the information contained in the classical receptive field. However, it is known that neurons in the same layer can interact with each other through lateral interactions, for example in V1, and that this can be the basis for computational principles~\citep{chavane_revisiting_2022}. For example, the combination of neighboring orientations may contribute to image categorization~\citep{perrinet_edge_2015}. Furthermore, neural information may be modulated by feedback information, e.g. to distinguish a figure from its background~\citep{roelfsema_early_2016}. Feedback has been shown to be essential for building realistic models of primary visual areas~\citep{boutin_sparse_2020, boutin_effect_2020}, especially to explain non-linear mechanisms~\citep{boutin_pooling_2022}. Currently, mainly due to our use of convolutions, it is not possible to implement these recurrent connections in our implementation (lateral or feedback). However, by inserting new spikes into the list of spikes reaching presynaptic addresses, the generic model is able to incorporate them. While this is theoretically possible, it must be properly tuned in practice so that these recurrent connections do not amplify neuronal activity outside of a homeostatic state (by extinction or explosion).

Such recurrent activity would be essential for the implementation of predictive or anticipatory processes~\citep{Benvenuti2020}. This is essential in a neural system because it contains several different delays that require temporal alignment~\citep{hogendoorn_predictive_2019}. This has been modeled before to explain, for example, the flash-lag illusion~\citep{khoei_flash-lag_2017}. As mentioned previously, this could be implemented using generalized coordinates (i.e., variables such as position complemented by motion, acceleration, jerk,~\ldots), and  knowing that ``neurobiologically, using delay operators just means changing synaptic connection strengths to take different mixtures of generalized sensations and their prediction errors''~\citep{perrinet_active_2014}. Our proposed model using heterogeneous delays provides an alternative and elegant implementation solution to this problem.
%
\subsection{Perspectives}
%%%-----------------------------------------------------------------
%
In defining our task, we emphasized that the generation of events depends on the spatial gradient in each image. This gradient has both horizontal and vertical dimensions, and its maxima are generally orientation dependent. Taken together, these oriented edges form the contours of visual objects in the scene~\citep{koenderink_representation_1987}. Thus, there is an interdependence between motion information and orientation information within the event stream. It would be crucial to investigate this dependency further. This could be initiated by training the model on a dataset with labels that provide local orientation. Exploring this dependence will allow us to dissociate and integrate these two forms of visual information. In particular, it will allow us to consider that the definition of motion is more accurate perpendicular to an oriented contour (aka the aperture problem). Thus, it will allow us to implement recurrent prediction rules, such as those identified to dissociate this problem~\citep{perrinet_motion-based_2012}.

The model is trained on a low-level local motion detection task, and one might wonder if it could be trained on higher-level tasks. An example of such a task would be depth estimation in the visual scene. There are several sources of information for depth estimation, such as binocular disparity or changes in texture or shading, but in our case motion parallax would be the most important cue~\citep{rogers_motion_1979}. This is because objects that are close to an observer move relatively faster on the retina than an object that is far away, and also because visual occlusions depend on depth order. Using this information, it is possible to segment objects and estimate their depth~\citep{yoonessi_contribution_2011}. However, this would first require the computation of the optic flow, i.e., the extension of the framework described here for a rigid full-field motion to a more general one where the motion may vary in the visual field. One possible implementation is to add a new layer to our model, analogous to the hierarchical organization which is prevalent in the visual cortex. This is theoretically possible by using the output of our model (which estimates motion in retinotopic space) as input to a new layer of neurons that would estimate motion in the visual field, including the depth dimension in the output supervision labels. This could have direct and important applications, e.g. in autonomous driving to detect obstacles in a fast and robust way. Another extension would be to actively generate sensor motion (physical or virtual) to obtain better depth estimates, especially to disambiguate uncertain estimates~\citep{nawrot_eye_2003}.
